---
title: "Intro to R Tidyverse - Notebook"
output: html_notebook
---
```{r}
library(dplyr)
install.packages("tidyverse")


x <- 4 #numerical
# x = 4

y <- TRUE 
#logical
z <- "Kevin" 
#character

x

x*2

x+4

mode(mtcars)
class(mtcars)

mode(iris)
class(iris)

library(magrittr)

install.packages("dplyr")

# explore dataset

head(iris) #first values
tail(iris) #last values
names(iris) #names of columns and rows
dim(iris) #dimensions
summary(iris) #stats of dataset
colnames(iris) #column names

summary(tail(iris))

iris %>% summary
# pipe operator: output da primeira funcao serve de input para segunda funcao


iris %>%  head(5)

iris %>%  head(5) %>% summary %>% class


library(dplyr)

#dplyr - hadley wickham
#important commands filter, arrange, select

filter(iris)

arrange(head(iris, 10))

Setosa = filter(iris, Sepal.Width>3 || filter(iris, Species=="setosa"))

filter(iris, Species!="setosa",Sepal.Width>3)

#ver com o pipe operator







Setosa

head(mtcars)

select(mtcars, 1:4)

select(mtcars, 1:4,am)

select(mtcars, -am) #minus/except column am

select(mtcars, starts_with("m")) #column starting with

select(mtcars, ends_with("q"))

select(mtcars, contains("q"))

select(mtcars, contains("q"), 1:3) #precedence order

iris2 = group_by(iris, Species) #same as iris %>% group_by(Species)
mode(iris2)

class(iris2)

#tdl_df tbl for dev

summarize(iris2, mean(Sepal.Width))

iris2

summarize(iris2, sd(Sepal.Width))

#median()
#sd() ->std dev
#var() ->variance

mtcars %>% group_by(vs,am) %>% summarize(mean(disp))

airquality

airquality %>% colnames

airquality %>% summary

airquality %>% group_by(Month) %>% summarize(mean(Ozone))
#NAs - no values

#how NAs affect the dataset
x = c(1, 2, 10, NA)
mean(x)

help(mean)

mean(x, na.rm = TRUE)

airquality %>% group_by(Month) %>% summarize(mean(Ozone, na.rm=TRUE))

#filter + select -> row + column
#group_by + summarize

arrange(mtcars, am)
#sort by am

#two level of ordering arrange(mtcars, am, vs)
#invert arrange(mtcars, -am)
#invert arrange(mtcars, desc(am))


#calculations
#add column based on other values
mutate(mtcars, ratio = mpg/wt)

mt2 = mutate(mtcars, ratio = mpg/wt)

help(mutate)

help(select)

select(mt2, -mpg, -wt, -ratio, everything())
#ordenar com os valores mpg rt e ratio no fim



```
summarise can be useful to give a quick preview of the data, particularly when the datasets are large
```{r}
library(ggplot2)

data(diamonds)

dim(diamonds)

head(diamonds)

diamonds %>% arrange(-price, cut) %>% select(price, cut) %>% arrange(-price)

```
using summarise to find mean price of diamonds in order of another of their properties (carat, cut, color...)
```{r}

library(dplyr)
diamonds %>% summary
# to find the headers of this dataset - carat, cut, color, clarity
# ex TODO: which property is a better predictor of price?

diamonds %>% select(carat, price) %>% summary

MeanPrice = summarise(group_by(diamonds, carat), mean(price))
MeanPrice
```
maybe summarize carat values - group into intervals
```{r}
help("group_by")

```


organize/arrange table by mean price - descending
```{r}

MeanPrice %>% arrange(desc(Mean))

```
T test
```{r}
data(smiths)

x = 1:99

t.test(x)

mean(x)
ttest=t.test(x, mu=4)

class(ttest)
mode(ttest)
names(ttest)

summary(ttest)

ttest$p.value

mtcars$disp

names(iris)

iris$Species

table(iris$Species) #freq table

table(iris$Species, iris$Petal.Width) #cross tabulation

#pearson chi squared test
chisq.test(mtcars$am, mtcars$vs)

library(broom)

install.packages("broom")

library(broom) #tidy up data!

tidy(ttest)

1:9

sample(1:9, 2) #x rolls of the dice, numbers 1 to 9
#it's a shuffler, never gives repeated values


#if rolls of the dice are larger than length...
sample(1:3, 4)

#fixed by...
sample(1:3, 4, replace=TRUE)

sum(sample(1:3, 4, replace=TRUE))

x1 <- 1:20
x2 <- sample(1:6, 20, replace=TRUE)
x3 <- c("A", "B", "C", "D")

x1
x2
x3

x4 <- rep(x3) #represent
x4

x4 <- rep(x3, each=2)
x4

x4 <- rep(x3, 5)
x4



x5 <-rnorm(20,100,10) #nr of results, mean, std dev
x5

x5 <- round(rnorm(20,100,10))
x5

x5 > 105 #evaluates if true or false

x6 <- x5 > 105
mean(x6) #mean of true and false, mean of zeros and ones...

#to find out how many are TRUE (x6 <- x5 > 105)...
sum(x6)

dataframe1 <- data.frame(ID = 101:120, x1, x2, x3)
dataframe2 <- data.frame(ID = 104:123, x4, x5, x6)
dataframe2
dataframe1

#join two tables (with conflicting IDs)
help(join)

inner_join(dataframe1, dataframe2) #intersection - only rows with matching keys in both x and y

left_join(dataframe1, dataframe2) #all rows in x plus y rows with matching keys in both x and y

full_join(dataframe1, dataframe2) #all of it

anti_join(dataframe1, dataframe2)#reverse full join, diagnose what's empty

anti_join(dataframe2, dataframe1) #what's left behind from df2
```





https://github.com/RWorkshop/workshopdatasets/blob/master/cappun.csv dataset url

Choose Raw version of csv values

urlfile<-'https://github.com/RWorkshop/workshopdatasets/blob/master/cappun.csv'
idaho<-read.csv(url(urlfile)) to import dataset

```{r}
urlfile<-'https://raw.githubusercontent.com/RWorkshop/workshopdatasets/master/cappun.csv'
cappun<-read.csv(url(urlfile))
```
```{r}
library(dplyr)
cappun %>% names 
cappun$age %>% mean
```
Grammar Graphics Plotting
```{r}

library(ggplot2)

cappun$age %>% plot
```

Let's graph some real data

aes() is aesthetics 
So far we're only setting up our canvas

```{r}
canvas = ggplot(aes(x=Sepal.Length, y=Sepal.Width), data=iris)
canvas
```
turn this into a scatter plot
```{r}
canvas+geom_point()
```
```{r}
help(group)
```


New chunk - graph experiment w/ boxplots
```{r}
help("geom_boxplot")
canvas+
  geom_boxplot(aes(group = cut_width(Sepal.Length, 7)), varwidth = TRUE)
```
RGB colours, default max value is 1, so to use the 255 scale change to maxColorValue = 255, output value is a hex 
```{r}
help(rgb)
testcor <- rgb(1, 0, 100, names = paste("testcolor"), maxColorValue = 255)
testcor

```

group = cut_width, cut_interval, cut_number... how to "cut" or group groups of data

```{r}
test <- rgb(170, 100, 100, names = paste("testcolor"), maxColorValue = 255)
test

canvas+
  geom_boxplot(aes(group = cut_interval(Sepal.Length, 7)), outlier.shape = 5, outlier.colour = test, fill = test)
```


Histogram (single variable, x)
```{r}
canvas2 = ggplot(aes(x=Sepal.Length), data=iris)
canvas2+geom_histogram()
```
	
`stat_bin()` using `bins = 30`. Pick better value with `binwidth`.

```{r}
help("stat_bin")
canvasingle = ggplot(aes(x=Sepal.Length), data=iris)
canvasingle+geom_histogram(binwidth = )
```


geom needs to suit the data or it won't work

```{r}
canvas2+geom_histogram()
+facet_wrap("Species")
```
Facet Wrap allows us to divide the display by different categories
```{r}
cars = ggplot(aes(x=mpg, y=wt), data=mtcars)
cars+geom_point()+facet_wrap("vs")+stat_smooth()
```
More graphical methods
```{r}
cars+geom_point()+facet_wrap("vs")+stat_smooth(method = "lm")
```
A plot is made of multiple layers
Layers consist of data, a set of mappings btwn variables and (?), a geometric object and statistical transformations

Add a trend line

```{r}
cars = ggplot(aes(x=mpg, y=wt), data=mtcars)
cars+geom_line()
```


```{r}
carcanvas = ggplot(aes(x=wt, y=mpg, colour=factor(am)), data=mtcars)
carcanvas+geom_line()+facet_wrap("vs")+stat_smooth(method = "lm")
```


univariate
bivariate
multivariate

```{r}
dcanvas = ggplot(aes(x=carat, y=price, colour=factor(cut)), data=diamonds)
dcanvas+geom_point()+stat_smooth()
```

```{r}
dcanvas = ggplot(aes(x=carat, y=price, colour=cut), data=diamonds)
dcanvas+geom_point()
```
Themes
```{r}
help("theme")
dcanvas+geom_point()+theme_minimal()
```



















